---
title: "The Wisdom of the Fantasy Crowd"
subtitle: "Does averaging mediocre projections beat picking one good source?"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
ffa_db <- readRDS(db_path("ffa_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))

# see 01-source-accuracy.qmd: ~38% of nfl_players_points rows have pts == NA
# (tracked but did not play) and must be dropped before any error metric.
actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

src <- ffa_db$ffa_proj_source_points |>
  filter_final() |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId))

joined <- src |> inner_join(actual, by = c("playerId", "season", "week"))

report_match_rate(
  nrow(distinct(joined, season, week, playerId)),
  nrow(distinct(src, season, week, playerId)),
  "wisdom-of-crowd bridge match rate"
)

RELIABLE_MIN_N <- 200
reliable_sources <- joined |>
  count(data_src) |>
  filter(n >= RELIABLE_MIN_N) |>
  pull(data_src)

overlap <- ffa_db$ffa_proj_source_points |>
  filter(tag == "final") |>
  distinct(season, week, id, data_src) |>
  count(season, week, id, name = "n_sources")
```

## A data-shape finding first

The natural "wisdom of the crowd" test would average each source's
*simultaneous* projection for the same player-week and see if the average
beats any individual source. That test needs several sources to actually
cover the same player-week at the same time. In this dataset's
`ffa_proj_source_points` table, at `tag == "final"`, **`r scales::percent(mean(overlap$n_sources > 1))` of
player-weeks (`r sum(overlap$n_sources > 1)` of `r nrow(overlap)`) have more
than one source recorded** — essentially every player-week has exactly one
`data_src` row. The 11 sources were not all scraped in parallel for the same
game across most of 2020-2025; coverage is fragmented by season instead (see
[01-source-accuracy](01-source-accuracy.md)'s season table). So a row-level
ensemble of *this* table isn't buildable at meaningful scale — we pivot to
the only genuine multi-source consensus the data does contain:
`ffa_projtable`'s precomputed `avg_type %in% c("average", "robust")`
columns, built upstream from a wider scrape than what survived into
`ffa_proj_source_points`.

## Does the ready-made consensus beat a single source?

```{r consensus-vs-source}
by_source_reliable <- joined |>
  filter(data_src %in% reliable_sources) |>
  group_by(method = data_src) |>
  reframe(error_metrics(pts, points)) |>
  mutate(kind = "individual source")

consensus <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type %in% c("average", "robust")) |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  inner_join(actual, by = c("playerId", "season", "week")) |>
  group_by(method = avg_type) |>
  reframe(error_metrics(pts, points)) |>
  mutate(kind = "ffa_projtable consensus", method = paste("ffa_projtable", method))

all_methods <- bind_rows(by_source_reliable, consensus)

all_methods |>
  ggplot(aes(mae, fct_reorder(method, mae, .desc = TRUE), color = kind)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("individual source" = "#1b9e77", "ffa_projtable consensus" = "#d95f02")) +
  labs(title = "Precomputed consensus vs. individual reliable sources", x = "MAE (points)", y = NULL, color = NULL) +
  theme_reports()
```

```{r consensus-by-pos}
consensus_by_pos <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type %in% c("average", "robust")) |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  inner_join(actual, by = c("playerId", "season", "week")) |>
  group_by(avg_type, pos) |>
  reframe(error_metrics(pts, points))

consensus_by_pos |>
  ggplot(aes(pos, mae, fill = avg_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(average = "#1b9e77", robust = "#d95f02")) +
  labs(title = "average vs. robust consensus, by position", x = NULL, y = "MAE (points)", fill = NULL) +
  theme_reports()
```

```{r table}
all_methods |> arrange(mae) |> mutate(across(where(is.numeric), \(x) round(x, 2)))
```

**Takeaway:** check whether either `ffa_projtable` consensus column beats
the best reliable individual source from
[01-source-accuracy](01-source-accuracy.md). If it does, that's real
evidence for "wisdom of crowds" even though we can't reconstruct which raw
sources fed it row-by-row in this dataset. If not, the apparent benefit of
"consensus" in fantasy football media may just be inheriting whichever
source dominates the input scrape (CBS, by volume — see report 01), not
genuine error reduction from averaging.
