---
title: "Who Is Actually Predictable?"
subtitle: "Separating talent, production and risk"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
ffa_db <- readRDS(db_path("ffa_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))

actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  only_games()

consensus <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type == "average") |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  only_games()

joined <- consensus |>
  inner_join(actual, by = c("playerId", "season", "week")) |>
  mutate(surprise = pts - points) |>
  left_join(dplyr::select(nfl_players_db$nfl_players, playerId, name), by = "playerId")

report_match_rate(
  nrow(distinct(joined, season, week, playerId)),
  nrow(distinct(consensus, season, week, playerId)),
  "predictability bridge match rate"
)

MIN_GAMES <- 6
```

## Executive summary

Instead of "who scores the most," we ask "who do we actually know how to
project?" `predictability = SD(actual - projected)` per player (lower =
more predictable), computed only for players with at least `r MIN_GAMES`
matched games. Plotting average points against predictability separates
four archetypes: reliable stars, boom/bust stars, low-upside-but-steady
players, and low-upside-and-volatile players.

```{r reliability-map}
reliability <- joined |>
  group_by(playerId, name, pos) |>
  summarise(
    games = n(),
    avg_pts = mean(pts),
    predictability = sd(surprise),
    .groups = "drop"
  ) |>
  filter(games >= MIN_GAMES)

med_pts <- median(reliability$avg_pts)
med_pred <- median(reliability$predictability)

reliability |>
  ggplot(aes(predictability, avg_pts, size = games, color = pos)) +
  geom_vline(xintercept = med_pred, linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = med_pts, linetype = "dashed", color = "grey60") +
  geom_point(alpha = 0.5) +
  scale_color_position() +
  labs(
    title = "Fantasy reliability map",
    subtitle = "Low predictability (left) = more consistent with the projection",
    x = "Predictability = SD(actual − projected)", y = "Average actual points", size = "Games", color = NULL
  ) +
  theme_reports()
```

```{r quadrant-table}
reliability |>
  mutate(
    quadrant = case_when(
      avg_pts >= med_pts & predictability < med_pred ~ "Reliable star",
      avg_pts >= med_pts & predictability >= med_pred ~ "Boom/bust star",
      avg_pts < med_pts & predictability < med_pred ~ "Consistent floor",
      TRUE ~ "Volatile low-upside"
    )
  ) |>
  count(quadrant, pos) |>
  ggplot(aes(pos, n, fill = quadrant)) +
  geom_col(position = "dodge") +
  labs(title = "Player archetypes by position", x = NULL, y = "Players", fill = NULL) +
  theme_reports()
```

```{r most-least}
bind_rows(
  reliability |> filter(avg_pts >= med_pts) |> slice_min(predictability, n = 10) |> mutate(kind = "most predictable star"),
  reliability |> filter(avg_pts >= med_pts) |> slice_max(predictability, n = 10) |> mutate(kind = "least predictable star")
) |>
  mutate(label = paste0(name, " (", pos, ")")) |>
  ggplot(aes(predictability, fct_reorder(label, predictability), color = kind)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("most predictable star" = "#1b9e77", "least predictable star" = "#d95f02")) +
  labs(title = "Most and least predictable above-average scorers", x = "Predictability (lower = more consistent)", y = NULL, color = NULL) +
  theme_reports()
```

**Takeaway:** the top-left quadrant (reliable stars) is who you draft early
with confidence; the top-right (boom/bust stars) is who wins or loses your
week depending on variance, not skill at projecting them.
