---
title: "Does Projection Uncertainty Mean Anything?"
subtitle: "Checking sd_pts and floor/ceiling against what actually happened"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
ffa_db <- readRDS(db_path("ffa_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))

actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  only_games()

consensus <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type == "average") |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  only_games()

joined <- consensus |>
  inner_join(actual, by = c("playerId", "season", "week")) |>
  filter(!is.na(sd_pts), !is.na(floor), !is.na(ceiling))

report_match_rate(
  nrow(distinct(joined, season, week, playerId)),
  nrow(distinct(consensus, season, week, playerId)),
  "uncertainty-calibration bridge match rate (incl. sd_pts/floor/ceiling non-NA)"
)
```

## Executive summary

`ffa_projtable` carries `sd_pts`, `floor` and `ceiling` alongside every
projection. If those numbers mean anything, "high uncertainty" players
should actually show larger errors, and `actual` should land inside
`[floor, ceiling]` a stable, sensible share of the time. `floor`/`ceiling`
aren't documented anywhere as a fixed nominal interval (e.g. not stated to
be an 80% or 95% band), so this report reports the *observed* coverage
rather than testing against an assumed target.

```{r calibration-by-tercile}
by_tercile <- joined |>
  mutate(uncertainty = ntile(sd_pts, 3),
         uncertainty = factor(uncertainty, labels = c("low", "medium", "high"))) |>
  group_by(uncertainty) |>
  reframe(error_metrics(pts, points))

by_tercile |>
  ggplot(aes(uncertainty, mae, fill = uncertainty)) +
  geom_col() +
  scale_fill_manual(values = c(low = "#1b9e77", medium = "#66a61e", high = "#d95f02")) +
  labs(title = "Realized MAE by projected-uncertainty tercile", x = "sd_pts tercile", y = "MAE (points)", fill = NULL) +
  theme_reports()
```

## Interval coverage

```{r coverage}
coverage <- joined |>
  mutate(inside = pts >= floor & pts <= ceiling)

overall_coverage <- mean(coverage$inside)

coverage_by_pos <- coverage |>
  group_by(pos) |>
  summarise(coverage = mean(inside), n = n(), .groups = "drop")

coverage_by_pos |>
  ggplot(aes(pos, coverage, fill = pos)) +
  geom_col() +
  geom_hline(yintercept = overall_coverage, linetype = "dashed", color = "grey40") +
  scale_fill_position() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of actuals landing inside [floor, ceiling]",
    subtitle = paste0("Dashed line = overall coverage (", scales::percent(overall_coverage, accuracy = 0.1), ")"),
    x = NULL, y = "Coverage", fill = NULL
  ) +
  theme_reports()
```

```{r coverage-by-tercile}
joined |>
  mutate(uncertainty = ntile(sd_pts, 3),
         uncertainty = factor(uncertainty, labels = c("low", "medium", "high")),
         inside = pts >= floor & pts <= ceiling) |>
  group_by(uncertainty) |>
  summarise(coverage = mean(inside), n = n(), .groups = "drop")
```

**Takeaway:** if MAE climbs cleanly from the low to the high `sd_pts`
tercile, the model's self-reported uncertainty is doing real work, not just
decoration. If coverage is roughly flat across positions and tercile
groups, `floor`/`ceiling` behaves like a consistent interval rather than one
that's only calibrated for some positions.
