---
title: "The Cost of a Questionable Tag"
subtitle: "How injury status relates to projected and actual production"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
ffa_db <- readRDS(db_path("ffa_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))

actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  only_games()

consensus <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type == "average") |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  only_games()

# nfl_player_injury_status has no season/week, only a weekly-snapshot
# timestamp. assign_week() maps each snapshot to the nearest preceding
# week boundary using ffa_projtable(final)'s own timestamps as the
# calendar -- an approximation, not the official NFL schedule. Excludes
# week == 0 (a season-total/ROS row, not a weekly calendar boundary).
calendar <- ffa_db$ffa_projtable |> filter_final() |> mutate(season = as.integer(season), week = as.integer(week)) |> only_games()

injury <- nfl_players_db$nfl_player_injury_status |>
  bind_cols(assign_week(nfl_players_db$nfl_player_injury_status$timestamp, calendar)) |>
  filter(!is.na(season)) |>
  mutate(
    status = case_match(
      injuryGameStatus,
      NA ~ "Healthy",
      "Questionable" ~ "Questionable",
      "Doubtful" ~ "Doubtful",
      "Out" ~ "Out",
      .default = "Other inactive"
    ),
    status = factor(status, levels = c("Healthy", "Questionable", "Doubtful", "Out", "Other inactive"))
  )
```

## Executive summary

`nfl_player_injury_status` only carries a `playerId` + weekly snapshot
`timestamp`, no `season`/`week` -- we derive those by nearest-preceding-week
lookup against `ffa_projtable`'s own timestamp calendar (see `assign_week()`
in `R/_common.R`); treat week assignment as approximate, not the official
schedule. `injuryGameStatus` has 12 raw levels; we bucket them into
Healthy (no note) / Questionable / Doubtful / Out / Other inactive (IR, PUP,
suspended, etc.). We compare same-week projected vs. actual points by
bucket, and check whether the projection already discounts questionable
players before the game is played.

```{r join-week}
weekly_status <- injury |>
  distinct(playerId, season, week, status) |>
  # a player can carry more than one status note in the same week window;
  # keep the most severe for that week
  mutate(severity = as.integer(status)) |>
  group_by(playerId, season, week) |>
  slice_max(severity, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(-severity)

by_status <- weekly_status |>
  inner_join(consensus, by = c("playerId", "season", "week")) |>
  inner_join(actual, by = c("playerId", "season", "week"))

report_match_rate(
  nrow(distinct(by_status, playerId, season, week)),
  nrow(distinct(weekly_status, playerId, season, week)),
  "injury-status to projection/actual match rate"
)
```

```{r cost-of-tag}
summary_by_status <- by_status |>
  group_by(status) |>
  summarise(
    n = n(),
    mean_projected = mean(points),
    mean_actual = mean(pts),
    .groups = "drop"
  )

summary_by_status |>
  pivot_longer(c(mean_projected, mean_actual), names_to = "kind", values_to = "value") |>
  ggplot(aes(value, status, color = kind)) +
  geom_line(aes(group = status), color = "grey70") +
  geom_point(size = 3) +
  scale_color_manual(values = c(mean_projected = "#1b9e77", mean_actual = "#d95f02"),
                      labels = c(mean_projected = "Projected", mean_actual = "Actual")) +
  labs(title = "The cost of an injury tag", x = "Mean points", y = NULL, color = NULL) +
  theme_reports()
```

## Do projections already discount questionable players?

```{r discount-check}
healthy_baseline <- summary_by_status$mean_projected[summary_by_status$status == "Healthy"]

summary_by_status |>
  mutate(pct_of_healthy_projection = mean_projected / healthy_baseline) |>
  select(status, n, mean_projected, mean_actual, pct_of_healthy_projection) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

```{r next-week}
# actual is already only_games()-filtered, so a week==1 status row's
# week - 1 lookup correctly finds no match (there is no week 0 game) instead
# of accidentally landing on the week==0 season-total row.
next_week_actual <- actual |> mutate(week = week - 1) |> rename(pts_next_week = pts)

by_status |>
  left_join(next_week_actual, by = c("playerId", "season", "week")) |>
  filter(!is.na(pts_next_week)) |>
  group_by(status) |>
  summarise(mean_next_week = mean(pts_next_week), n = n(), .groups = "drop") |>
  ggplot(aes(status, mean_next_week, fill = status)) +
  geom_col() +
  labs(title = "Next-week production after this week's injury tag", x = NULL, y = "Mean points, following week", fill = NULL) +
  theme_reports() +
  theme(legend.position = "none")
```

**Takeaway:** if `pct_of_healthy_projection` drops well below 1.0 for
Questionable/Doubtful, the projection pipeline is already pricing in the
injury note before kickoff. If `mean_actual` still falls far short of even
that discounted projection, the market isn't discounting enough. Watch for
a selection-bias trap here, though: "Healthy" absorbs every player who
simply never got a status note logged, including irrelevant bench/waiver
players with low baseline points -- so a *higher* mean for Questionable
than Healthy more likely means "only notable, heavily-used players get
injury notes at all" than "questionable players outscore healthy ones."
Comparing status buckets within the same player (their own healthy-week
average vs. their questionable-week average) would remove this confound;
this report doesn't do that decomposition.
