---
title: "Which Position Is Hardest to Predict?"
subtitle: "Decomposing projection error by position"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
ffa_db <- readRDS(db_path("ffa_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))

actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

consensus <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type == "average") |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId))

joined <- consensus |>
  inner_join(actual, by = c("playerId", "season", "week")) |>
  mutate(error = pts - points)

report_match_rate(
  nrow(distinct(joined, season, week, playerId)),
  nrow(distinct(consensus, season, week, playerId)),
  "error-by-position bridge match rate"
)
```

## Executive summary

Using the same `ffa_projtable` consensus (`avg_type == "average"`) vs.
actual points as [03-booms-busts](03-booms-busts.md), we break the error
down by position to see which one the projection model consistently
struggles with most.

```{r distribution}
joined |>
  ggplot(aes(error, pos, fill = pos)) +
  geom_violin(alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_position() +
  labs(title = "Distribution of projection error by position", x = "Actual − projected (points)", y = NULL, fill = NULL) +
  theme_reports() +
  theme(legend.position = "none")
```

```{r ecdf}
joined |>
  ggplot(aes(abs(error), color = pos)) +
  stat_ecdf() +
  scale_color_position() +
  labs(title = "ECDF of absolute error by position", x = "|Actual − projected| (points)", y = "Cumulative share", color = NULL) +
  theme_reports()
```

```{r ranking}
by_pos <- joined |>
  group_by(pos) |>
  reframe(error_metrics(pts, points)) |>
  arrange(desc(mae))

by_pos |>
  ggplot(aes(mae, fct_reorder(pos, mae))) +
  geom_segment(aes(xend = 0, yend = pos), color = "grey70") +
  geom_point(aes(color = pos), size = 4) +
  scale_color_position() +
  labs(title = "Hardest position to predict (by MAE)", x = "MAE (points)", y = NULL) +
  theme_reports() +
  theme(legend.position = "none")
```

```{r table}
by_pos |> mutate(across(where(is.numeric), \(x) round(x, 2)))
```

**Takeaway:** high-usage skill positions with more scoring pathways
(receiving + rushing, or big-play upside) tend to have both higher average
points and higher absolute error -- the ranking above shows whether that
scales roughly with scoring volume or whether one position is
disproportionately hard to call relative to how many points is at stake.
