---
title: "The Shifting Shape of Fantasy Scoring"
subtitle: "Positional production, 2020-2025"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))

points <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

players <- nfl_players_db$nfl_players |>
  select(playerId, position) |>
  clean_positions(pos_col = position)

joined <- points |> inner_join(players, by = "playerId")

report_match_rate(nrow(joined), nrow(points), "positional-evolution player-metadata match rate")
```

## Executive summary

Average fantasy production by position, season and week, 2020-2025 (this
dataset's coverage -- not back to 2019). We look at whether RBs have lost
ground to WRs and whether scoring has concentrated in fewer positions or
spread out.

```{r stacked-share}
# week == 0 is the precomputed season-total row -- use it directly instead
# of summing every weekly row (which would double the total).
by_season_pos <- joined |>
  filter(week == 0) |>
  group_by(season, pos = position) |>
  summarise(total_pts = sum(pts), .groups = "drop")

by_season_pos |>
  ggplot(aes(season, total_pts, fill = pos)) +
  geom_area(position = "fill") +
  scale_fill_position() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Share of total fantasy points by position", x = "Season", y = "Share of total points", fill = NULL) +
  theme_reports()
```

```{r avg-trend}
# per-game average, so week == 0 (season total) must be excluded here even
# though the previous chunk uses it for total_pts.
by_season_pos_avg <- joined |>
  only_games() |>
  group_by(season, pos = position) |>
  summarise(avg_pts = mean(pts), .groups = "drop")

by_season_pos_avg |>
  ggplot(aes(season, avg_pts, color = pos)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_color_position() +
  labs(title = "Average points per player-game by position", x = "Season", y = "Average points", color = NULL) +
  theme_reports()
```

```{r rb-wr}
by_season_pos_avg |>
  filter(pos %in% c("RB", "WR")) |>
  pivot_wider(names_from = pos, values_from = avg_pts) |>
  mutate(wr_minus_rb = WR - RB) |>
  ggplot(aes(season, wr_minus_rb)) +
  geom_col(fill = "#7570b3") +
  geom_hline(yintercept = 0, color = "grey40") +
  labs(title = "WR average points − RB average points, by season", x = "Season", y = "Difference (points)") +
  theme_reports()
```

**Takeaway:** a rising WR-minus-RB gap over the seasons is data-level
confirmation of the "Zero RB" drafting theory; a flat or shrinking gap
would argue against it for this league's scoring settings.
