---
title: "Stars, Starters and Replacement Players"
subtitle: "Empirical Value Over Replacement by position"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))

points <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

players <- nfl_players_db$nfl_players |>
  select(playerId, name, position) |>
  clean_positions(pos_col = position)

season_totals <- points |>
  inner_join(players, by = "playerId") |>
  group_by(season, playerId, name, pos = position) |>
  summarise(season_pts = sum(pts), .groups = "drop") |>
  group_by(season, pos) |>
  mutate(rank = rank(-season_pts, ties.method = "first")) |>
  ungroup()
```

## Executive summary

This computes VOR **empirically** from actual season point totals, not from
`ffa_projtable$points_vor` -- that column is 100% NA for 2020-2023 and only
appears in 2024-2025, so it can't support a full-history VOR story.
Replacement rank per position (QB14/RB35/WR40/TE14, K/DST extended to 14)
is a fixed convention from the source brief, **not** derived from this
league's actual starter demand: `nfl_teams_rosters$slotPosition` only has 3
coarse buckets (`O`/`DT`/`K`), too coarse to reconstruct real per-position
starter counts, so this is a documented simplification, not a league-exact
number.

```{r vor}
replacement_pts <- season_totals |>
  filter(rank == REPLACEMENT_RANK[pos]) |>
  select(season, pos, replacement_pts = season_pts)

vor <- season_totals |>
  left_join(replacement_pts, by = c("season", "pos")) |>
  mutate(vor = season_pts - replacement_pts)

vor |>
  filter(rank <= 60) |>
  ggplot(aes(rank, season_pts, color = pos)) +
  geom_line(aes(group = interaction(season, pos)), alpha = 0.15) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  geom_vline(data = tibble(pos = names(REPLACEMENT_RANK), rank = REPLACEMENT_RANK),
             aes(xintercept = rank, color = pos), linetype = "dashed", alpha = 0.4) +
  scale_color_position() +
  labs(
    title = "Positional scarcity curve",
    subtitle = "Dashed line = replacement rank; thin lines = individual seasons, bold = average",
    x = "Positional rank within season", y = "Season total points", color = NULL
  ) +
  theme_reports()
```

```{r top-vor}
vor |>
  filter(season == max(season)) |>
  group_by(pos) |>
  slice_max(vor, n = 5) |>
  ungroup() |>
  mutate(label = paste0(name, " (", pos, ")")) |>
  ggplot(aes(vor, fct_reorder(label, vor), fill = pos)) +
  geom_col() +
  scale_fill_position() +
  labs(title = paste("Top VOR by position,", max(vor$season)), x = "VOR (points above replacement)", y = NULL, fill = NULL) +
  theme_reports() +
  theme(legend.position = "none")
```

**Takeaway:** the steepness of the scarcity curve right up to the
replacement-rank line is "positional scarcity" made visible -- a position
whose curve drops fast even at low ranks creates more competitive advantage
per roster spot than one with a long, flat tail of replaceable production.
