---
title: "Was Your Team Good, or Just Lucky?"
subtitle: "All-play record vs. actual record, 2023-2025"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_round_db <- readRDS(db_path("nfl_round_db"))
nfl_teams_db <- readRDS(db_path("nfl_teams_db"))

# nfl_teams_week_stats only has tag == "final" for 2023-2025; 2020-2022
# only exists as tag == "preview" (no finalized actual score). This report
# is scoped to 2023-2025 for that reason, not by choice.
team_scores <- nfl_round_db$nfl_teams_week_stats |>
  filter(tag == "final", statId == "pts") |>
  mutate(season = as.integer(season), week = as.integer(week), score = as.numeric(value))

matchups <- nfl_round_db$matchups_games |>
  filter(awayTeamOutcome != "", homeTeamOutcome != "") |>
  mutate(season = as.integer(season), week = as.integer(week))

report_match_rate(
  n_distinct(team_scores$season),
  3,
  "schedule-luck season coverage (expect 3: 2023-2025)"
)
```

## Executive summary

`nfl_teams_week_stats` only has a finalized (`tag == "final"`) actual score
per team-week for 2023-2025; 2020-2022 only exists as a preview snapshot,
so this report is scoped to 2023-2025. For every team-week we compute the
"all-play" record -- how many of the other teams that team would have beaten
with that same score -- and compare season-total all-play wins to actual
wins.

```{r all-play}
all_play <- team_scores |>
  group_by(season, week) |>
  mutate(all_play_wins_this_week = rank(score, ties.method = "average") - 1) |>
  ungroup()

actual_wins <- bind_rows(
  matchups |> filter(awayTeamOutcome == "win") |> transmute(season, week, teamId = awayTeamTeamId, won = 1),
  matchups |> filter(homeTeamOutcome == "win") |> transmute(season, week, teamId = homeTeamTeamId, won = 1)
)

luck <- all_play |>
  left_join(actual_wins, by = c("season", "week", "teamId")) |>
  mutate(won = coalesce(won, 0)) |>
  group_by(season, teamId) |>
  summarise(
    actual_wins = sum(won),
    all_play_wins = sum(all_play_wins_this_week),
    weeks = n(),
    .groups = "drop"
  ) |>
  left_join(select(nfl_teams_db$nfl_teams, teamId, name), by = "teamId") |>
  mutate(luck = actual_wins - all_play_wins)
```

```{r scatter}
luck |>
  ggplot(aes(all_play_wins, actual_wins, color = factor(season))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Actual wins vs. all-play wins",
    subtitle = "Above the line = lucky, below = unlucky",
    x = "All-play wins (season total)", y = "Actual wins", color = "Season"
  ) +
  theme_reports()
```

```{r luck-ranking}
luck |>
  mutate(label = paste0(name, " (", season, ")")) |>
  ggplot(aes(luck, fct_reorder(label, luck), fill = luck > 0)) +
  geom_col() +
  geom_vline(xintercept = 0, color = "grey40") +
  scale_fill_manual(values = c(`TRUE` = "#1b9e77", `FALSE` = "#d95f02"), guide = "none") +
  labs(title = "Luckiest and unluckiest team-seasons", x = "Actual wins − all-play wins", y = NULL) +
  theme_reports() +
  theme(axis.text.y = element_text(size = 7))
```

**Takeaway:** a team well above the diagonal won more than its weekly
scores deserved -- schedule luck, not roster strength. That's the team to
regress toward the mean next season; the team below the diagonal is the
buy-low candidate.
