---
title: "The Manager Effect"
subtitle: "Who leaves points on the bench?"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_round_db <- readRDS(db_path("nfl_round_db"))
nfl_teams_db <- readRDS(db_path("nfl_teams_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))

actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

# nfl_teams_rosters keeps multiple timestamps even within tag == "final"
# (snapshots through the week); latest_snapshot() collapses to the single
# final lineup before counting starters/bench.
rosters <- nfl_round_db$nfl_teams_rosters |>
  filter_final() |>
  latest_snapshot(c("season", "week", "teamId")) |>
  mutate(is_starter = rosterSlotId %in% STARTER_SLOTS) |>
  filter(rosterSlotId %in% c(STARTER_SLOTS, BENCH_SLOTS)) |>
  inner_join(actual, by = c("playerId", "season", "week"))

report_match_rate(nrow(rosters), nrow(latest_snapshot(filter_final(nfl_round_db$nfl_teams_rosters), c("season","week","teamId"))), "roster-to-actual-points match rate")
```

## Executive summary

`actual` = sum of points scored by a team's starters
(`rosterSlotId %in% STARTER_SLOTS`, 7 lineup slots/week).
`possible` = sum of the top-K point scorers among that team's **entire**
roster that week (K = number of starter slots), ignoring exact
position-eligibility (QB/RB/WR/TE/FLEX/K/DST). This is a **documented
simplification**, not a real lineup optimizer -- `nfl_teams_rosters$slotPosition`
only has 3 coarse buckets (`O`/`DT`/`K`), too coarse to reconstruct which
bench player was actually eligible to start at which slot, so `possible`
is a modest overestimate of what a manager could really have started.
`efficiency = actual / possible` is still a useful relative ranking across
managers even though its absolute level isn't exact.

```{r efficiency}
team_week <- rosters |>
  group_by(season, week, teamId) |>
  summarise(
    n_starters = sum(is_starter),
    actual = sum(pts[is_starter]),
    possible = sum(sort(pts, decreasing = TRUE)[seq_len(sum(is_starter))]),
    .groups = "drop"
  ) |>
  filter(n_starters > 0, possible > 0) |>
  mutate(efficiency = actual / possible)

team_season <- team_week |>
  group_by(season, teamId) |>
  summarise(efficiency = mean(efficiency), weeks = n(), .groups = "drop") |>
  left_join(select(nfl_teams_db$nfl_teams, teamId, name), by = "teamId")

team_season |>
  mutate(label = paste0(name, " (", season, ")")) |>
  ggplot(aes(efficiency, fct_reorder(label, efficiency))) +
  geom_col(fill = "#1b9e77") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Points captured / possible, by team-season", x = "Efficiency", y = NULL) +
  theme_reports() +
  theme(axis.text.y = element_text(size = 7))
```

```{r worst-weeks}
team_week |>
  left_join(select(nfl_teams_db$nfl_teams, teamId, name), by = "teamId") |>
  mutate(left_on_bench = possible - actual) |>
  slice_max(left_on_bench, n = 15) |>
  mutate(label = paste0(name, " (", season, " wk", week, ")")) |>
  ggplot(aes(left_on_bench, fct_reorder(label, left_on_bench))) +
  geom_col(fill = "#d95f02") +
  labs(title = "Worst single-week lineup decisions", x = "Points left on the bench (approx.)", y = NULL) +
  theme_reports() +
  theme(axis.text.y = element_text(size = 7))
```

**Takeaway:** treat `efficiency` as a relative ranking, not an exact
percentage -- the top managers are consistently starting their
highest-scoring roster spots, while the bottom of the list is where
lineup-setting discipline (or waiver-wire activity) is actually costing
wins.
