---
title: "Did the Model Know?"
subtitle: "Reliability of the Monte Carlo simulation quantiles"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
sim_db <- readRDS(db_path("dudes_simulation_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))

actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

# dudes_players_simulations$simType has 18 distinct values but coverage is
# fragmented by season -- most are single-season tags (<=125 rows) rather
# than real simulation methodologies. proj_src and current_season_his are
# the only two with meaningful, overlapping coverage (both 2023-2024), so
# this report is scoped to those two and to those seasons -- no 2025 data.
sims <- sim_db$dudes_players_simulations |>
  filter(simType %in% c("proj_src", "current_season_his")) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  select(season, week, playerId, pos, simType, simQuantiles) |>
  mutate(.row = row_number()) |>
  unnest(simQuantiles) |>
  pivot_wider(names_from = name, values_from = value) |>
  select(-.row)

joined <- sims |> inner_join(actual, by = c("playerId", "season", "week"))

report_match_rate(nrow(joined), nrow(sims), "simulation-to-actual match rate")
```

## Executive summary

For each simulated player-week we have quantiles at 5/15/30/50/70/85/95%. If
the simulation is well-calibrated, the actual result should fall below the
50% quantile about half the time, inside `[15%, 85%]` about 70% of the
time, and inside `[5%, 95%]` about 90% of the time. This report is scoped
to `simType %in% c("proj_src", "current_season_his")` and seasons
2023-2024 -- the only simulation methodology with real, overlapping
coverage in this dataset (see note above).

```{r reliability}
thresholds <- tribble(
  ~label, ~nominal,
  "below q50", 0.50,
  "inside [q15, q85]", 0.70,
  "inside [q05, q95]", 0.90
)

reliability <- joined |>
  group_by(simType) |>
  summarise(
    n = n(),
    `below q50` = mean(pts <= `50%`),
    `inside [q15, q85]` = mean(pts >= `15%` & pts <= `85%`),
    `inside [q05, q95]` = mean(pts >= `5%` & pts <= `95%`),
    .groups = "drop"
  ) |>
  pivot_longer(-c(simType, n), names_to = "label", values_to = "observed") |>
  left_join(select(thresholds, label, nominal), by = "label")

reliability |>
  ggplot(aes(nominal, observed, color = simType)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 4) +
  geom_text(aes(label = label), vjust = -1, size = 3, show.legend = FALSE) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Simulation calibration: predicted vs. observed coverage",
    subtitle = "On the diagonal = well-calibrated",
    x = "Nominal probability", y = "Observed proportion", color = NULL
  ) +
  theme_reports()
```

```{r table}
reliability |> mutate(across(where(is.numeric), \(x) round(x, 3)))
```

**Takeaway:** points above the diagonal mean the simulation is
under-confident (actual results land inside the interval more often than
promised) -- safe but not sharp. Points below mean it's over-confident,
which is the riskier failure mode for anyone making decisions off these
quantiles.
