---
title: "A Fantasy Football Periodic Table"
subtitle: "Clustering players by production, volatility and availability"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))

points <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

players <- nfl_players_db$nfl_players |>
  select(playerId, name, position) |>
  clean_positions(pos_col = position)

MIN_GAMES <- 6

features <- points |>
  inner_join(players, by = "playerId") |>
  group_by(season, playerId, name, pos = position) |>
  summarise(
    games = n(),
    production = mean(pts),
    volatility = sd(pts),
    .groups = "drop"
  ) |>
  filter(games >= MIN_GAMES, !is.na(volatility)) |>
  mutate(cv = volatility / pmax(production, 0.1))

report_match_rate(nrow(features), nrow(distinct(points, season, playerId)), "periodic-table eligible player-seasons (>= min games)")
```

## Executive summary

Per-player-season features: production (mean points), volatility (SD of
points), CV (volatility / production) and availability (games played),
restricted to players with at least `r MIN_GAMES` games. `kmeans()` (base R,
scaled features) groups players into archetypes, which we label after the
fact from each cluster's centroid -- the labels are descriptive, not
categories that exist in the source data.

```{r cluster}
set.seed(42)
X <- features |> transmute(production, cv, games) |> scale()
km <- kmeans(X, centers = 5, nstart = 25)
features$cluster <- factor(km$cluster)

centroids <- features |>
  group_by(cluster) |>
  summarise(production = mean(production), cv = mean(cv), games = mean(games), n = n(), .groups = "drop") |>
  arrange(desc(production))

centroids
```

```{r label}
# Labels derived from each cluster's centroid position (high/low production
# x high/low cv), not a fixed mapping -- read off centroids above.
labels <- centroids |>
  mutate(
    label = case_when(
      production >= median(features$production) & cv <= median(features$cv) ~ "Reliable star",
      production >= median(features$production) & cv > median(features$cv) ~ "Boom/bust star",
      production < median(features$production) & cv <= median(features$cv) ~ "Safe floor",
      TRUE ~ "Bench lottery"
    )
  ) |>
  select(cluster, label)

features <- features |> left_join(labels, by = "cluster")
```

```{r scatter}
features |>
  ggplot(aes(cv, production, color = label)) +
  geom_point(alpha = 0.5) +
  labs(title = "Fantasy Football periodic table", x = "Coefficient of variation (volatility / production)", y = "Production (avg points)", color = "Archetype") +
  theme_reports()
```

```{r counts}
features |>
  count(label, pos) |>
  ggplot(aes(pos, n, fill = label)) +
  geom_col(position = "dodge") +
  labs(title = "Archetype mix by position", x = NULL, y = "Player-seasons", fill = NULL) +
  theme_reports()
```

**Takeaway:** "Reliable star" is who you draft to win your league with the
least stress; "Boom/bust star" wins or loses your week for you; "Bench
lottery" is where late-round upside picks live.
