---
title: "Career Fingerprints"
subtitle: "Weekly production heatmaps for the top scorers, 2020-2025"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))

points <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week))

players <- nfl_players_db$nfl_players |> select(playerId, name)

joined <- points |> inner_join(players, by = "playerId")

TOP_N <- 10

top_players <- joined |>
  group_by(playerId, name) |>
  summarise(career_pts = sum(pts), .groups = "drop") |>
  slice_max(career_pts, n = TOP_N)
```

## Executive summary

Weekly fantasy points for the top `r TOP_N` all-time scorers in this
dataset's coverage (2020-2025, not back to 2019). Color intensity =
points; gaps = bye weeks, injuries or seasons not played.

```{r heatmap}
joined |>
  semi_join(top_players, by = "playerId") |>
  mutate(name = factor(name, levels = top_players$name)) |>
  ggplot(aes(week, season, fill = pts)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma") +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  facet_wrap(~name, ncol = 2) +
  labs(title = "Career fingerprint: weekly points by season", x = "Week", y = "Season", fill = "Points") +
  theme_reports() +
  theme(panel.grid = element_blank())
```

```{r ranking}
top_players |>
  ggplot(aes(career_pts, fct_reorder(name, career_pts))) +
  geom_col(fill = "#1b9e77") +
  labs(title = paste("Top", TOP_N, "all-time scorers (2020-2025)"), x = "Career points", y = NULL) +
  theme_reports()
```

**Takeaway:** a consistently bright row is a durable, healthy star; a row
with a dark gap mid-career marks an injury or decline that's easy to miss
in a season-average stat line but obvious here.
