---
title: "Does FGW Persist?"
subtitle: "Testing whether trailing FGW predicts next weeks better than points or projections (2023-2025)"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_round_db <- readRDS(db_path("nfl_round_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))
ffa_db <- readRDS(db_path("ffa_db"))

# Same construction as 15-fgw-metrics.qmd, scoped to 2023-2025 for the same
# reason (finalized team scores only exist for those 3 seasons).
team_scores <- team_week_scores(nfl_round_db)
outcomes <- team_week_outcomes(nfl_round_db)
score_outcome <- team_scores |> inner_join(outcomes, by = c("season", "week", "teamId"))
win_prob <- win_prob_model(score_outcome)
team_sd <- sd(score_outcome$team_score)

players_pos <- nfl_players_db$nfl_players |> select(playerId, pos = position) |> clean_positions(pos_col = pos)
rosters <- nfl_round_db$nfl_teams_rosters |>
  filter_final() |>
  latest_snapshot(c("season", "week", "teamId")) |>
  filter(rosterSlotId %in% STARTER_SLOTS) |>
  mutate(season = as.integer(season), week = as.integer(week))
actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  only_games()
starters <- rosters |> inner_join(actual, by = c("playerId", "season", "week")) |> inner_join(players_pos, by = "playerId")
baseline_pos <- starters |> group_by(season, pos) |> summarise(baseline = mean(pts), .groups = "drop")

proj <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type == "average") |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  only_games() |>
  distinct(season, week, playerId, .keep_all = TRUE) |>
  select(season, week, playerId, proj_points = points)

fgw <- starters |>
  left_join(baseline_pos, by = c("season", "pos")) |>
  left_join(team_scores, by = c("season", "week", "teamId")) |>
  left_join(proj, by = c("season", "week", "playerId")) |>
  mutate(
    delta = pts - baseline,
    fgw_classic = pnorm(delta / team_sd) - 0.5,
    efgw = win_prob(team_score) - win_prob(team_score - delta)
  )

report_match_rate(sum(!is.na(fgw$proj_points)), nrow(fgw), "starter-week-to-projection match rate (for the trailing-projection baseline)")
```

## Executive summary

`fgw_analysis.md` section (Experiment 2) asks a narrower question than
[15-fgw-metrics](15-fgw-metrics.md): not "does FGW describe what already
happened," but **"does a player's FGW through week *N* predict their
production in weeks *N+1..N+4* better than trailing points-per-game, or the
projection itself?"** If it doesn't, FGW/eFGW are retrospective-only tools
-- genuinely useful for 15's attribution question, but not a forecasting
signal. We test this the same way [04-predictability](04-predictability.md)
frames reliability: rolling trailing windows within each of the 3 seasons,
requiring at least 3 prior starts and comparing against at least 1 of the
next 4 weeks.

```{r rolling}
CUTOFFS <- 4:13 # leaves a 4-week-or-shorter future window inside a 17-week season
HORIZON <- 4
MIN_TRAIL <- 3

roll <- map_dfr(CUTOFFS, function(N) {
  trailing <- fgw |>
    filter(week <= N) |>
    group_by(season, playerId, pos) |>
    summarise(
      n_trail = n(), trailing_ppg = mean(pts), trailing_fgw = mean(fgw_classic),
      trailing_efgw = mean(efgw), trailing_proj = mean(proj_points, na.rm = TRUE),
      .groups = "drop"
    ) |>
    filter(n_trail >= MIN_TRAIL)

  future <- fgw |>
    filter(week > N, week <= N + HORIZON) |>
    group_by(season, playerId) |>
    summarise(n_future = n(), future_ppg = mean(pts), .groups = "drop")

  trailing |> inner_join(future, by = c("season", "playerId")) |> mutate(cutoff_week = N)
})
```

Each row is one (player, season, cutoff week) observation: a trailing
feature computed only from weeks up to the cutoff, and a future outcome
computed only from weeks strictly after it -- no look-ahead.
`r nrow(roll)` such observations across `r n_distinct(roll$playerId)`
starters and the 3 seasons.

```{r persistence-corr}
persistence <- tibble(
  predictor = c("Trailing PPG", "Trailing projection", "Trailing FGW-classic", "Trailing eFGW"),
  corr_with_future = c(
    cor(roll$trailing_ppg, roll$future_ppg),
    cor(roll$trailing_proj, roll$future_ppg, use = "complete.obs"),
    cor(roll$trailing_fgw, roll$future_ppg),
    cor(roll$trailing_efgw, roll$future_ppg)
  )
)

persistence |>
  ggplot(aes(corr_with_future, fct_reorder(predictor, corr_with_future))) +
  geom_segment(aes(xend = 0, yend = predictor), color = "grey70") +
  geom_point(size = 4, color = "#1b9e77") +
  labs(
    title = "Correlation with next 1-4 weeks' actual PPG",
    subtitle = "Higher = better forecast of what a player does next",
    x = "Correlation with future PPG", y = NULL
  ) +
  theme_reports()
```

```{r table}
persistence |> arrange(desc(corr_with_future)) |> mutate(corr_with_future = round(corr_with_future, 3))
```

**Takeaway:** trailing projection and trailing PPG both predict a player's
next few weeks meaningfully better than trailing FGW-classic or eFGW do.
That's the expected result, not a disappointing one -- `fgw_analysis.md`
itself (sections 8 and 19) frames FGW as a retrospective attribution metric,
not a forecasting one: it deliberately compresses a player's raw production
through their team's win-probability curve, which discards exactly the kind
of information (this player is simply good/healthy/featured) that carries
forward week to week. FGW answers "how much did this performance matter,"
not "what will this player do next" -- for the latter, keep using the
projection. This also means a decision rule built by summing *past* FGW
would be a weaker predictor than one built from projected points --
[17-fgw-decision-backtest](17-fgw-decision-backtest.md) uses projections
(xFGW), not trailing FGW, for exactly this reason.
