---
title: "Would xFGW Have Won You More Games?"
subtitle: "A historical start/sit policy backtest: Points vs. VOR vs. xFGW (2023-2025)"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

```{r load}
nfl_round_db <- readRDS(db_path("nfl_round_db"))
nfl_stats_db <- readRDS(db_path("nfl_stats_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))
ffa_db <- readRDS(db_path("ffa_db"))

team_scores <- team_week_scores(nfl_round_db)
outcomes <- team_week_outcomes(nfl_round_db)
score_outcome <- team_scores |> inner_join(outcomes, by = c("season", "week", "teamId"))
win_prob <- win_prob_model(score_outcome)

players_pos <- nfl_players_db$nfl_players |> select(playerId, pos = position) |> clean_positions(pos_col = pos)
actual <- nfl_stats_db$nfl_players_points |>
  filter(!is.na(pts)) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  only_games()
proj <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type == "average") |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  only_games() |>
  distinct(season, week, playerId, .keep_all = TRUE) |>
  select(season, week, playerId, proj_points = points)
```

## Executive summary

[15-fgw-metrics](15-fgw-metrics.md) and
[16-fgw-persistence](16-fgw-persistence.md) test whether FGW describes and
predicts production. This report runs `fgw_analysis.md`'s central test
(Experiment 3, sections 10-13): **for every real historical start/sit
decision this league actually faced -- a starter and the real bench players
eligible to replace them at that slot -- would ranking by projected points,
by VOR (points above position replacement), or by xFGW (win-probability
shift) have picked someone different, and would that pick have won more
games?**

Bench alternatives are reconstructed from `rosterSlotId` (confirmed to
resolve cleanly to QB/RB/WR/TE/K/DST/bench, with slot 5 = FLEX pooling
RB/WR/TE), not the coarser `slotPosition` bucket
[11-manager-effect](11-manager-effect.md) had to work around.

```{r decision-pool}
all_roster <- nfl_round_db$nfl_teams_rosters |>
  filter_final() |>
  latest_snapshot(c("season", "week", "teamId")) |>
  filter(rosterSlotId %in% c(STARTER_SLOTS, 20)) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  inner_join(players_pos, by = "playerId") |>
  left_join(actual, by = c("playerId", "season", "week")) |>
  left_join(proj, by = c("playerId", "season", "week")) |>
  mutate(pts = coalesce(pts, 0))

baseline_proj <- all_roster |>
  filter(rosterSlotId %in% STARTER_SLOTS, !is.na(proj_points)) |>
  group_by(season, pos) |> summarise(baseline = mean(proj_points), .groups = "drop")
team_proj_hist <- all_roster |>
  filter(rosterSlotId %in% STARTER_SLOTS, !is.na(proj_points)) |>
  group_by(season, week, teamId) |> summarise(team_proj = sum(proj_points), .groups = "drop")
team_actual_hist <- all_roster |>
  filter(rosterSlotId %in% STARTER_SLOTS) |>
  group_by(season, week, teamId) |> summarise(team_actual = sum(pts), .groups = "drop")

starters <- all_roster |>
  filter(rosterSlotId %in% STARTER_SLOTS, !is.na(proj_points)) |>
  mutate(rowid = row_number(), req_pos = SLOT_POSITION[as.character(rosterSlotId)])
bench <- all_roster |> filter(rosterSlotId == 20, !is.na(proj_points))

decisions <- starters |>
  left_join(team_proj_hist, by = c("season", "week", "teamId")) |>
  left_join(team_actual_hist, by = c("season", "week", "teamId"))

report_match_rate(nrow(decisions), nrow(starters), "starter decisions with team-projection context (should be ~100%, same join)")

# FLEX (slot 5) pools RB/WR/TE bench candidates; every other slot maps to
# exactly one position via SLOT_POSITION.
decisions_exp <- decisions |>
  mutate(elig_pos = ifelse(rosterSlotId == FLEX_SLOT, list(FLEX_ELIGIBLE), as.list(req_pos))) |>
  unnest(elig_pos)

candidates <- decisions_exp |>
  select(rowid, season, week, teamId, elig_pos) |>
  inner_join(
    bench |> select(season, week, teamId, elig_pos = pos, cand_playerId = playerId, cand_proj = proj_points, cand_pts = pts),
    by = c("season", "week", "teamId", "elig_pos"), relationship = "many-to-many"
  )
```

For each decision, the "pool" is the actual starter plus every real,
same-slot-eligible bench player on that team that week, each with a
hypothetical team total if they'd been swapped in for the incumbent --
using **projected** points (for the Points/VOR/xFGW policy decision, made
before the game) and **actual** points (to score what really would have
happened).

```{r pool}
pool <- bind_rows(
  decisions |> transmute(rowid, season, week, teamId, playerId, proj = proj_points, pts, pos,
                          team_proj, team_actual, incumbent_proj = proj_points, incumbent_pts = pts, is_incumbent = TRUE),
  candidates |> transmute(rowid, playerId = cand_playerId, proj = cand_proj, pts = cand_pts, pos = elig_pos) |>
    left_join(decisions |> select(rowid, season, week, teamId, team_proj, team_actual, incumbent_proj = proj_points, incumbent_pts = pts), by = "rowid") |>
    mutate(is_incumbent = FALSE)
) |>
  left_join(baseline_proj, by = c("season", "pos")) |>
  mutate(
    delta = proj - baseline,
    hyp_team_proj = team_proj - incumbent_proj + proj, # this slot swapped, rest of team held at its real projection
    hyp_win_prob = win_prob(hyp_team_proj),
    hyp_team_actual = team_actual - incumbent_pts + pts # same swap, using real results
  )

picks <- pool |>
  group_by(rowid) |>
  summarise(
    points_pick = playerId[which.max(proj)], points_actual = hyp_team_actual[which.max(proj)],
    vor_pick = playerId[which.max(delta)], vor_actual = hyp_team_actual[which.max(delta)],
    xfgw_pick = playerId[which.max(hyp_win_prob)], xfgw_actual = hyp_team_actual[which.max(hyp_win_prob)],
    incumbent = playerId[is_incumbent],
    n_options = n(),
    .groups = "drop"
  ) |>
  left_join(select(decisions, rowid, season, week, teamId), by = "rowid") |>
  left_join(select(score_outcome, season, week, teamId, historical_win = won), by = c("season", "week", "teamId")) |>
  left_join(
    score_outcome |> select(season, week, teamId = opponentId, opp_score = team_score) |> distinct(),
    by = c("season", "week", "teamId")
  ) |>
  filter(n_options > 1, !is.na(opp_score)) |> # a real decision needs >=1 alternative and a known opponent score
  mutate(
    points_win = as.integer(points_actual > opp_score),
    vor_win = as.integer(vor_actual > opp_score),
    xfgw_win = as.integer(xfgw_actual > opp_score),
    flip_vs_points = xfgw_pick != points_pick
  )
```

## Decision Flip Rate

```{r flip-rate}
flip_rate <- mean(picks$flip_vs_points)
n_decisions <- nrow(picks)
n_flipped <- sum(picks$flip_vs_points)
```

Across `r n_decisions` real start/sit decisions with at least one eligible
bench alternative, xFGW disagreed with ranking by projected points in
**`r n_flipped`** of them -- a flip rate of
**`r scales::percent(flip_rate, accuracy = 0.1)`**. `fgw_analysis.md`
section 12 frames this as the test to run *before* looking at outcomes: a
low flip rate means the sophistication mostly doesn't change anything in
practice, whatever the theory says.

## Did the flip help?

```{r win-rate-table}
tibble(
  situation = c("All decisions", "Decisions where xFGW and Points differ"),
  n = c(nrow(picks), n_flipped),
  historical = c(mean(picks$historical_win), mean(picks$historical_win[picks$flip_vs_points])),
  points_policy = c(mean(picks$points_win), mean(picks$points_win[picks$flip_vs_points])),
  vor_policy = c(mean(picks$vor_win), mean(picks$vor_win[picks$flip_vs_points])),
  xfgw_policy = c(mean(picks$xfgw_win), mean(picks$xfgw_win[picks$flip_vs_points]))
) |>
  mutate(across(c(historical, points_policy, vor_policy, xfgw_policy), \(x) scales::percent(x, accuracy = 0.1)))
```

`historical` is what the real manager's actual lineup that week produced
(win rate ≈ 50% everywhere, by construction of a win/loss league). The
other three columns are the hypothetical team win rate if *every* decision
in that row had instead followed that policy, holding every other slot at
its real, historical result.

## Calibration: does the win-probability model itself hold up ex-ante?

```{r calibration}
teamweek_cal <- decisions |>
  distinct(season, week, teamId, team_proj) |>
  left_join(select(score_outcome, season, week, teamId, won), by = c("season", "week", "teamId")) |>
  filter(!is.na(won)) |>
  mutate(pred = win_prob(team_proj))

brier <- mean((teamweek_cal$pred - teamweek_cal$won)^2)

teamweek_cal |>
  mutate(bin = ntile(pred, 5)) |>
  group_by(bin) |>
  summarise(mean_predicted = mean(pred), observed_win_rate = mean(won), n = n(), .groups = "drop") |>
  ggplot(aes(mean_predicted, observed_win_rate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 4, color = "#1b9e77") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Calibration: win_prob(team's projected total) vs. real outcome",
    subtitle = paste0("Brier score = ", round(brier, 3), " (0.25 = always-guess-50%, lower is better); quintile bins, ~108 team-weeks each"),
    x = "Predicted P(win), quintile mean", y = "Observed win rate"
  ) +
  theme_reports()
```

This is a harder test than [15-fgw-metrics](15-fgw-metrics.md)' curve: that
one was fit and evaluated on the same real, *actual* scores. Here the same
`win_prob()` curve is fed each team's **projected** total instead --
genuinely ex-ante, using only information available before the games were
played.

**Takeaway:** the flip rate is low (`r scales::percent(flip_rate, accuracy = 0.1)`)
-- most start/sit calls simply don't have a same-slot alternative close
enough for the win-probability framing to matter, echoing
`fgw_analysis.md` section 12's "sofisticamos a matemática e quase nunca
mudamos a decisão" scenario. And on the `r n_flipped` decisions where it
*did* disagree with Points -- small enough that this is not a statistically
powered test -- xFGW's hypothetical win rate did not come out ahead of
simply starting the higher-projected player. Calibration is real but modest
(Brier score clearly better than the 0.25 always-guess-50% baseline, but
short of a sharp forecaster). Taken together, this 3-season sample gives no
evidence that xFGW would have won this league more games than projected
points already do -- consistent with `fgw_analysis.md` section 2's own
warning and section 19's own expectation, not a failure of the metric's
construction. The honest reading of the doc's H0/H1 framing: **H0 is not
rejected here.**
