---
title: "Is a Waiver Add Worth More Than Its VOR?"
subtitle: "A single retrospective RosterValue case study from real simulation data (2024)"
---

```{r setup}
source("R/_common.R")
library(tidyverse)
library(dm)
```

## Executive summary

`fgw_analysis.md`'s Experiment 4 asks whether a `RosterValue(p) = Σ [P(Win |
Roster + p) − P(Win | Roster)]` rule would pick better waiver adds than
simply ranking by VOR. Of the four building blocks this FGW series depends
on, Monte Carlo player simulations are by far the weakest: `simType %in%
c("proj_src", "current_season_his")` in `dudes_players_simulations` only
covers **14 season-weeks total** (2023 weeks 1-11, 2024 weeks 2-4) and
**nothing at all in 2025**. A real backtest the way
[17-fgw-decision-backtest](17-fgw-decision-backtest.md) runs one isn't
possible on that little data, so this report is scoped down to what
`fgw_analysis.md`'s own framing already anticipates for the weakest-evidence
experiment: **one concrete, real, retrospective case study**, not a general
claim. Treat this report's conclusion as illustrative, not as evidence at
the same confidence level as 15-17.

```{r load}
nfl_round_db <- readRDS(db_path("nfl_round_db"))
nfl_players_db <- readRDS(db_path("nfl_players_db"))
ffa_db <- readRDS(db_path("ffa_db"))
sim_db <- readRDS(db_path("dudes_simulation_db"))

# simQuantiles is a 7-row (name, value) tibble per player-week -- same shape
# 12-simulation-calibration.qmd unnests. Build a wide q05..q95 table once.
sim_q <- sim_db$dudes_players_simulations |>
  filter(simType %in% c("proj_src", "current_season_his")) |>
  mutate(season = as.integer(season), week = as.integer(week)) |>
  select(season, week, playerId, simType, simQuantiles) |>
  # the two simType values overlap on only 13 of 6575 player-weeks;
  # current_season_his has denser coverage per 12-simulation-calibration.qmd
  arrange(season, week, playerId, simType == "current_season_his") |>
  distinct(season, week, playerId, .keep_all = TRUE) |>
  mutate(.row = row_number()) |>
  unnest(simQuantiles) |>
  pivot_wider(names_from = name, values_from = value) |>
  select(-.row, -simType)

PROBS <- c(0.05, 0.15, 0.30, 0.50, 0.70, 0.85, 0.95)
QCOLS <- c("5%", "15%", "30%", "50%", "70%", "85%", "95%")

# Linear-interpolate a player-week's 7 quantile points into a resampleable
# inverse-CDF -- the same idea as R/_common.R's win_prob_model(), applied to
# a single player's predictive score distribution instead of team outcomes.
sim_draw <- function(playerId_, season_, week_, n = 2000) {
  row <- sim_q |> filter(season == season_, week == week_, playerId == playerId_)
  if (nrow(row) == 0) return(rep(NA_real_, n))
  approxfun(PROBS, as.numeric(row[1, QCOLS]), rule = 2)(runif(n))
}
sim_median <- function(playerId_, season_, week_) {
  row <- sim_q |> filter(season == season_, week == week_, playerId == playerId_)
  if (nrow(row) == 0) return(NA_real_)
  as.numeric(row[1, "50%"])
}
```

## The case: a real RB2 decision, 2024 week 2

```{r case-setup}
SEASON <- 2024
DECISION_WEEK <- 2
HORIZON_WEEKS <- 3:4 # as far as proj_src/current_season_his coverage reaches from here
TARGET_TEAM <- 8
INCUMBENT <- 2570335 # Zach Charbonnet, the real RB2 starter that week

players_pos <- nfl_players_db$nfl_players |> select(playerId, pos = position) |> clean_positions(pos_col = pos)
proj <- ffa_db$ffa_projtable |>
  filter_final() |>
  filter(avg_type == "average") |>
  bridge_ffa_to_nfl(ffa_db$ffa_player_ids) |>
  clean_positions() |>
  filter(!is.na(playerId)) |>
  only_games() |>
  distinct(season, week, playerId, .keep_all = TRUE) |>
  select(season, week, playerId, proj_points = points)

rosters <- nfl_round_db$nfl_teams_rosters |>
  filter_final() |>
  latest_snapshot(c("season", "week", "teamId")) |>
  filter(season == SEASON, week == DECISION_WEEK, teamId == TARGET_TEAM, rosterSlotId %in% STARTER_SLOTS) |>
  inner_join(players_pos, by = "playerId")
other_starters <- rosters$playerId[rosters$playerId != INCUMBENT]

outcomes <- team_week_outcomes(nfl_round_db)
team_scores <- team_week_scores(nfl_round_db)
opp_scores <- outcomes |>
  filter(season == SEASON, teamId == TARGET_TEAM, week %in% HORIZON_WEEKS) |>
  left_join(team_scores |> rename(opponentId = teamId, opp_score = team_score), by = c("season", "week", "opponentId"))
```

Team `r TARGET_TEAM` started Zach Charbonnet at RB2 in week 2 of 2024, with
only Raheem Mostert as an actual same-position bench alternative. Weeks 3-4
are the horizon this simulation coverage reaches, and both were genuinely
tight team-projection-vs-opponent matchups (this team's projected total sat
within a few points of the opponent's real score both weeks) -- exactly the
kind of context where a marginal roster decision could plausibly matter,
unlike the blowout weeks that made most decisions in
[17-fgw-decision-backtest](17-fgw-decision-backtest.md) irrelevant to the
outcome regardless of who started.

```{r roster-value}
team_proj_excl <- map_dfr(opp_scores$week, function(w) {
  tibble(week = w, team_proj_excl = sum(
    proj$proj_points[proj$season == SEASON & proj$week == w & proj$playerId %in% other_starters], na.rm = TRUE
  ))
})

cov <- sim_q |> filter(season == SEASON, week %in% opp_scores$week) |> count(playerId) |> filter(n == nrow(opp_scores))

# candidate pool: any RB (this team's own bench, or any other team's bench --
# a real waiver-wire check for actual free-agent availability isn't done
# here, a documented simplification for this illustrative case) with
# simulation coverage across both horizon weeks
candidates <- nfl_round_db$nfl_teams_rosters |>
  filter_final() |>
  latest_snapshot(c("season", "week", "teamId")) |>
  filter(season == SEASON, week == DECISION_WEEK, rosterSlotId == 20) |>
  inner_join(players_pos, by = "playerId") |>
  filter(pos == "RB", playerId %in% cov$playerId) |>
  distinct(playerId)

pool <- unique(c(INCUMBENT, candidates$playerId))

roster_value <- map_dfr(pool, function(pid) {
  per_week <- map_dfr(seq_len(nrow(opp_scores)), function(i) {
    w <- opp_scores$week[i]
    opp <- opp_scores$opp_score[i]
    base <- team_proj_excl$team_proj_excl[team_proj_excl$week == w]
    draws <- sim_draw(pid, SEASON, w)
    tibble(p_win = mean(base + draws > opp, na.rm = TRUE), proj_med = sim_median(pid, SEASON, w))
  })
  tibble(playerId = pid, roster_value = mean(per_week$p_win), avg_proj = mean(per_week$proj_med, na.rm = TRUE))
})

roster_value <- roster_value |>
  left_join(select(nfl_players_db$nfl_players, playerId, name), by = "playerId") |>
  mutate(is_incumbent = playerId == INCUMBENT) |>
  arrange(desc(roster_value))
```

`RosterValue(p)` here is the average, across the 2 horizon weeks, of P(Team
+ candidate's simulated score > opponent's real score) − nothing subtracted
yet, since we're ranking candidates directly rather than reporting a lift.
`avg_proj` is each candidate's own simulated median score, averaged across
the same 2 weeks (`ffa_projtable` doesn't cover most bench RBs, so the
simulation's own median stands in as the "projected points" comparator).

```{r plot}
roster_value |>
  slice_max(roster_value, n = 15, with_ties = TRUE) |>
  mutate(label = if_else(is_incumbent, paste(name, "(started)"), name)) |>
  ggplot(aes(roster_value, fct_reorder(label, roster_value), fill = is_incumbent)) +
  geom_col() +
  scale_fill_manual(values = c(`TRUE` = "#1b9e77", `FALSE` = "#7570b3"), guide = "none") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "RosterValue by RB candidate (top 15)",
    subtitle = "P(win) if this player had occupied the RB2 slot, weeks 3-4 2024",
    x = "RosterValue", y = NULL
  ) +
  theme_reports()
```

```{r table}
roster_value |>
  mutate(roster_value = scales::percent(roster_value, accuracy = 0.1), avg_proj = round(avg_proj, 1)) |>
  select(name, avg_proj, roster_value, is_incumbent) |>
  head(10)
```

**Takeaway:** in this one case, RosterValue and the simulation's own
projected points rank candidates almost identically
(Spearman r ≈ 0.93 across all `r nrow(roster_value)` candidates) -- most of
RosterValue's ranking power is, again, just points in a different unit, the
same conclusion [15-fgw-metrics](15-fgw-metrics.md) and
[17-fgw-decision-backtest](17-fgw-decision-backtest.md) reached with the
empirical win-probability curve instead of simulation draws. What
RosterValue adds here is confirmation, not surprise: the real historical
start (Charbonnet) sits in the top tier of both rankings, and the team's
actual bench alternative (Mostert) sits at the very bottom -- the manager's
real decision checks out. That's a reassuring result for this one
team-week, not a general finding; with 2 usable horizon weeks, no 2025
simulation coverage at all, and one illustrative team picked specifically
because it was a close matchup, this report cannot and does not claim
RosterValue beats VOR at scale the way
[17-fgw-decision-backtest](17-fgw-decision-backtest.md) tested for
start/sit. A real answer to `fgw_analysis.md`'s Experiment 4 would need
either a broader simulation dataset than this league currently keeps, or
running the same `win_prob_model()`-based approach 15-17 use with
projections instead of simulations -- which this report intentionally
avoided, to keep at least one part of the FGW series grounded in the
doc's own "literally simulate it" framing (section 7) rather than another
normal-approximation shortcut.
